Verification of a TCP/IP stack

### 2021-05-12 SecDev21/paper.tex
Minor fixes

\author{%
\IEEEauthorblockN{Guillaume Cluzel}
\IEEEauthorblockA{\textit{Adacore \& ENS de Lyon}}
\and
\IEEEauthorblockN{Kyriakos Georgious}
\IEEEauthorblockA{\textit{Adacore \& University of Bristol}}
\and
\IEEEauthorblockN{Yannick Moy}
\IEEEauthorblockA{\textit{Adacore}}
\and
\IEEEauthorblockN{Clément Zeller}
\IEEEauthorblockA{\textit{Oryx Embedded}}
}

@@

Our approach in this work has been to incrementally evolve an existing TCP/IP
stack in C by replacing parts of the code with formally verified code in SPARK.
The stack chosen, CycloneTCP\footnote{\url{https://www.oryx-embedded.com/products/CycloneTCP.html}},
a mature TCP/IP stack targeted at embedded processors with no Operating System.
We present the TCP protocol in section~\ref{sec:TCP} and then briefly the
choosen stack, the Oryx Embedded one, in section~\ref{sec:stack}
before presenting the specification techniques used in section~\ref{sec:spec}
and then in sections~\ref{sec:verif} and~\ref{sec:API} we explain how we
formally implemented in SPARK the TCP user functions and how we hardened the
user's API. Finally we end in section~\ref{sec:results} by showing the results
of our work.

@@

The behavior of the three-phase communication can be described by the state
machine.
Every group of states corresponds to a one phase. More precisely \sclosed
stands for a closed connection. Both states \ssynsent and \ssynrcv denotes the
state that a TCP process can take during the initiation of a new connection. The
state \sestab represents the state when the connection is established,
\textit{i.e.} when two TCPs can communicate by sending and receiving data.
Then, two groups of states can be distinguished: the group formed by \sfwone,
\sfwtwo, \sclosing and \stimewait and that represents the state that a TCP can
reach when it is closing its connection and it is at the origin of the closing;
and the group formed by the state \sclosew and \slastack when the connection is
closing and the closing was initiated by the remote TCP.
Finally, the state \slisten stands for a TCP that is waiting for an incoming
connection.

@@

The socket structure is shared between different functions, and contains a lot
of fields. Most of the fields are used in other parts of library and are not
relevant for our verification. We only need to concentrate on a subset of fields
to make the specifications easier to write, and to understand.
This can be achieve in SPARK by using \emph{models} and \emph{ghost code}.
The ghost code is code uses by GNATprove to simplify the proof that is removed
by the compiler to keep the size of executable file reasonable.

@@

When a transition has to be made in the code, it is made through the helper
function \spark{TCP_Change_State}. The precondition of this function describes
all the allowed transitions. Therefore, we try to do an forbidden transition,
GNATprove will display a warning. This is the most direct way to make a
transition in the state machine.
As the TCP user functions are all protected with a mutex to avoid interferences
from other threads that could change the TCP state during their execution, in
particular when a new segment is received, it is quite easy to know the
possible set of state that can have the socket during the all execution of the
functions.

There is only one function that makes the verification harder because it
releases the mutex: the function \spark{TCP_Wait_For_Events}.
It takes as an arguments an event to wait and returns once the event
is triggered by the reception of a segment.
The role of this
function is to release the mutex and allow the reception task to process the
incoming segments. This function
can be used, for example, in the function \spark{open} when the connection
is initialized to wait until the connection has been established or not to
correctly set the error that indicates to the user the state of the connection.
If we consider that everything can happen during a call to \spark{TCP_Wait_For_Events},
the verification becomes too imprecise and it becomes impossible to prove the
functional specifications extracted from the norm. Instead we have decided to
find a better contract to this function.

@@

The function \spark{TCP_Wait_For_Events} first checks if the events is not already
true. If not, the mutex is released until a segment that triggered the event is
received or until the timeout is reach. Each time a new segment is received, it
is processed and if it contains a segment that causes a change of state, the TCP
state is updated. Then the events are updated, and if the waited event becomes
true due to the change of state, it is triggered.

@@

The TCP protocol implies a specific order such that the user can call the API's
functions without breaking the functional specification of the protocol. This
order also is being conveyed by the protocol's state machine given in
figure~\ref{Fig:statemachine}.
Ada's pre- and postconditions are a powerful tool to express such inter-function
dependencies, while SPARK technology can be used to guarantee that the
assertions always hold. Thus, post- and preconditions were introduced to model
a partial order on the calls to the API's functions.

@@

But these transition are not allowed by the norm, and thus not allowed by the
precondition of \spark{TCP_Change_State}. The problem is warm by SPARK and a fix
has been proposed.

>>>>

\author{%
\IEEEauthorblockN{Guillaume Cluzel}
\IEEEauthorblockA{\textit{AdaCore \& ENS de Lyon}}
\and
\IEEEauthorblockN{Kyriakos Georgious}
\IEEEauthorblockA{\textit{AdaCore \& University of Bristol}}
\and
\IEEEauthorblockN{Yannick Moy}
\IEEEauthorblockA{\textit{AdaCore}}
\and
\IEEEauthorblockN{Clément Zeller}
\IEEEauthorblockA{\textit{Oryx Embedded}}
}

@@

Our approach in this work has been to incrementally evolve an existing TCP/IP
stack in C by replacing parts of the code with formally verified code in SPARK.
The stack chosen, CycloneTCP\footnote{\url{https://www.oryx-embedded.com/products/CycloneTCP.html}},
a mature TCP/IP stack targeted at embedded processors with no Operating System.
We present the TCP protocol in section~\ref{sec:TCP} and then briefly the
chosen stack, the Oryx Embedded one, in section~\ref{sec:stack}
before presenting the specification techniques used in section~\ref{sec:spec}
and then in sections~\ref{sec:verif} and~\ref{sec:API} we explain how we
formally implemented in SPARK the TCP user functions and how we hardened the
user's API. Finally we end in section~\ref{sec:results} by showing the results
of our work.

@@

The behavior of the three-phase communication can be described by the state
machine.
Every group of states corresponds to a one phase. More precisely \sclosed
stands for a closed connection. Both states \ssynsent and \ssynrcv denote the
state that a TCP process can take during the initiation of a new connection. The
state \sestab represents the state when the connection is established,
\textit{i.e.} when two TCPs can communicate by sending and receiving data.
Then, two groups of states can be distinguished: the group formed by \sfwone,
\sfwtwo, \sclosing and \stimewait and that represents the state that a TCP can
reach when it is closing its connection and it is at the origin of the closing;
and the group formed by the state \sclosew and \slastack when the connection is
closing and the closing was initiated by the remote TCP.
Finally, the state \slisten stands for a TCP that is waiting for an incoming
connection.

@@

The socket structure is shared between different functions, and contains a lot
of fields. Most of the fields are used in other parts of library and are not
relevant for our verification. We only need to concentrate on a subset of fields
to make the specifications easier to write, and to understand.
This can be achieved in SPARK by using \emph{models} and \emph{ghost code}
which are used by GNATprove for analysis but not compiled in the final executable.

@@

When a transition has to be made in the code, it is made through the helper
function \spark{TCP_Change_State}. The precondition of this function describes
all the allowed transitions. Therefore, when we try to do a forbidden transition,
GNATprove will display a message. This is the most direct way to make a
transition in the state machine.
As the TCP user functions are all protected with a mutex to avoid interferences
from other threads that could change the TCP state during their execution, in
particular when a new segment is received, it is quite easy to know the
possible set of state that can have the socket during the all execution of the
functions.

There is only one function that makes the verification harder because it
releases the mutex: the function \spark{TCP_Wait_For_Events}.
It takes as argument an event to wait and returns once the event
is triggered by the reception of a segment.
The role of this
function is to release the mutex and allow the reception task to process the
incoming segments. This function
can be used, for example, in the function \spark{open} when the connection
is initialized to wait until the connection has been established or not to
correctly set the error that indicates to the user the state of the connection.
If we consider that everything can happen during a call to \spark{TCP_Wait_For_Events},
the verification becomes too imprecise and it becomes impossible to prove the
functional specifications extracted from the norm. Instead we have decided to
find a better contract to this function.

@@

The function \spark{TCP_Wait_For_Events} first checks if the event is not already
true. If not, the mutex is released until a segment that triggered the event is
received or until the timeout is reach. Each time a new segment is received, it
is processed and if it contains a segment that causes a change of state, the TCP
state is updated. Then the events are updated, and if the waited event becomes
true due to the change of state, it is triggered.

@@

The TCP protocol implies a specific order such that the user can call the API's
functions without breaking the functional specification of the protocol. This
order also is being conveyed by the protocol's state machine given in
figure~\ref{Fig:statemachine}.
Ada's pre- and postconditions are a powerful tool to express such inter-function
dependencies, while SPARK technology can be used to guarantee that the
assertions always hold. Thus, pre- and postconditions were introduced to model
a partial order on the calls to the API's functions.

@@

But these transition are not allowed by the norm, and thus not allowed by the
precondition of \spark{TCP_Change_State}. The problem is detected by SPARK and a fix
has been proposed.
